### 2016-07-27 Chapters/Ch03-PCMCoupledCluster.tex
Start on papers summaries and contributions
Some more work

%************************************************
\chapter{Continuum solvation models}\label{ch:CSM}
%************************************************
%*****************************************
%*****************************************
%*****************************************
%*****************************************
%*****************************************

>>>>

% mainfile: ../RobertoDiRemigioPhDThesis.tex
\chapter{Coupled cluster}\label{ch:PCMCC}

\section{Notation}
Cluster operator:
\begin{equation}
  T = \sum_{i=1}^\mathcal{M}\sum_{\mu_i}\tamp{\mu_i}\cluster{\mu_i}
\end{equation}
$\mu_i$ is the $\mu$-th excitation at the $i$-th
excitation level and $\mathcal{M}$ is the truncation level.

The term at the $j$-th excitation level in the left CC wavefunction:
\begin{equation}
  \bra{\tbar{j}} = \sum_{\mu_j}\tbar{\mu_j}\bra{\mu_j}
\end{equation}

\acl{MP} partitioning:
\begin{equation}
 H_0 = F + \Phi = \sum_P\epsilon_P\cons{P}\anni{P} + (g- V_\mathrm{HF})
\end{equation}
where $F$ if the Fock matrix and $\Phi$ is the fluctuation potential.
We assume to deal with an \acl{HF} (\acs{HF}) reference function.

Similarity transformed operator:
\begin{equation}
  \overline{O} =\expo{-T}O\expo{T}
\end{equation}
Since the Fock operator is diagonal, its similarity transformation
truncates after the second term:
\begin{equation}
  \overline{F} = F + \sum_{i}^4\sum_{\mu_i}\tamp{\mu_i}\cluster{\mu_i}\denom{\mu_i}
\end{equation}
where $\denom{\mu_i}$ is the difference in orbital energy between the
virtual and occupied spin-orbitals of excitation $i$.
For the fluctuation potential the similarity transformation truncates
after the fifth term, \ie{} includes up to four-fold nested commutators
with the cluster operator.

\acs{CC} expectation values are formed using the left and
right \acs{CC} states and are denoted as:
\begin{equation}
  O(\tamp{}, \tbar{}) = \braket{\mathrm{HF} | \overline{O} | \mathrm{HF} }
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{j} | \overline{O} | \mathrm{HF} }
\end{equation}
We will also use the notation $O^\mathcal{M}(\tamp{}, \tbar{})$ to
denote a \acs{CC} expectation value truncated at the $\mathcal{M}$-th
excitation level.

\section{Effective Coupled Cluster Lagrangian}

The effective CC Lagrangian can be readily obtained by coupling the
usual CC Lagrangian with the classical polarization functional.
Care must however be taken in the definition of the classical
variational degrees of freedom as to not erroneously include
contributions from the reference state in the subsequent correlated
treatment.

Let us assume that the reference state is the optimized single \acs{HF}
determinant in the presence of the implicit/explicit environment.
Then the corresponding variational degrees of freedom satisfy:
\begin{equation}
  \begin{pmatrix}
    \PCM & \MMPCM^\dagger \\
    \MMPCM & \MM
  \end{pmatrix}
  \begin{pmatrix}
   \sigma^\mathrm{HF} \\
   \kappa^\mathrm{HF}
  \end{pmatrix}
  =
  -
  \begin{pmatrix}
   \esp^\mathrm{HF} \\
   \zeta^\mathrm{HF}
  \end{pmatrix}
\end{equation}
and correspond to the \emph{reference reaction field}.
In the subsequent correlated treatment one can separate the reaction
field into reference and correlated components:
\begin{equation}
  \begin{pmatrix}
   \sigma \\
   \kappa
  \end{pmatrix}
  =
  \begin{pmatrix}
   \sigma^\mathrm{HF} \\
   \kappa^\mathrm{HF}
  \end{pmatrix}
  +
  \begin{pmatrix}
   \sigma^\mathrm{CC} \\
   \kappa^\mathrm{CC}
  \end{pmatrix}
\end{equation}
and similarly for the sources:
\begin{equation}
  \begin{pmatrix}
   \esp(\tamp{}, \tbar{}) \\
   \zeta(\tamp{}, \tbar{})
  \end{pmatrix}
  =
  \begin{pmatrix}
   \esp^\mathrm{HF} \\
   \zeta^\mathrm{HF}
  \end{pmatrix}
  +
  \begin{pmatrix}
    \tilde{\esp}(\tamp{}, \tbar{}) \\
   \tilde{\zeta}(\tamp{}, \tbar{})
  \end{pmatrix}
  =
  \begin{pmatrix}
  \braket{\mathrm{HF} | \esp  | \mathrm{HF}} \\
  \braket{\mathrm{HF} | \zeta | \mathrm{HF}}
  \end{pmatrix}
  +
  \begin{pmatrix}
   \braket{\mathrm{HF} | \tilde{\esp} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\esp} | \mathrm{HF}} \\
   \braket{\mathrm{HF} | \tilde{\zeta} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\zeta} | \mathrm{HF}}
  \end{pmatrix}
\end{equation}
The Baker--Campbell--Hausdorff expansions of the source terms have been
rewritten as:
\begin{equation}
  \overline{\s} = \s + \tilde{\s} = \s + \commutator{\s}{T}
  + \frac{1}{2}\commutator{\commutator{\s}{T}}{T}
\end{equation}
taking advantage of the fact that $\s$ is a one-electron operator and
hence its commutator expansion truncates at the third term.
We apply a shift to the correlation part of the source term
one-electron operator to remove the reference source terms:
\begin{equation}
  \begin{aligned}
  \begin{pmatrix}
   \braket{\mathrm{HF} | \tilde{\esp} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \tilde{\esp} | \mathrm{HF}} \\
   \braket{\mathrm{HF} | \tilde{\zeta} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \tilde{\zeta} | \mathrm{HF}}
  \end{pmatrix}
  &=
  \begin{pmatrix}
    \braket{\mathrm{HF} | \overline{\esp} - \esp^\mathrm{HF} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\esp} - \esp^\mathrm{HF} | \mathrm{HF}} \\
  \braket{\mathrm{HF} | \overline{\zeta} - \zeta^\mathrm{HF} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\zeta} -
  \zeta^\mathrm{HF}| \mathrm{HF}}
  \end{pmatrix}
  \\
  &=
  \begin{pmatrix}
   \braket{\mathrm{HF} | \tilde{\esp} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\esp} | \mathrm{HF}} \\
   \braket{\mathrm{HF} | \tilde{\zeta} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\zeta} | \mathrm{HF}}
  \end{pmatrix}
  =
  \begin{pmatrix}
    \tilde{\esp}(\tamp{}, \tbar{}) \\
    \tilde{\zeta}(\tamp{}, \tbar{})
  \end{pmatrix}
\end{aligned}
\end{equation}
which is equivalent to imposing normal ordering of the
operators.~\cite{Shavitt}
The total functional will include both reference and correlation
components:
\begin{equation}
  \begin{aligned}
  U_\mathrm{pol} &=
  \frac{1}{2}(\sigma^\mathrm{HF}+\sigma^\mathrm{CC})\PCM(\sigma^\mathrm{HF}+\sigma^\mathrm{CC})
+ (\sigma^\mathrm{HF}+\sigma^\mathrm{CC})(\esp^\mathrm{HF}+\tilde{\esp}(\tamp{},\tbar{})) \\
&+ \frac{1}{2}(\kappa^\mathrm{HF}+\kappa^\mathrm{CC})\MM(\kappa^\mathrm{HF}+\kappa^\mathrm{CC})
+ (\kappa^\mathrm{HF}+\kappa^\mathrm{CC})(\zeta^\mathrm{HF}+\tilde{\zeta}(\tamp{},\tbar{})) \\
&+ (\sigma^\mathrm{HF}+\sigma^\mathrm{CC})\bi{X}(\kappa^\mathrm{HF}+\kappa^\mathrm{CC}) \\
&=
    \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma^\mathrm{HF}
  + \sigma^\mathrm{HF}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa^\mathrm{HF}
  + \kappa^\mathrm{HF}\zeta^\mathrm{HF}
  + \sigma^\mathrm{HF}\bi{X}\kappa^\mathrm{HF} \\
&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{CC}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{CC}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{CC} \\
&+  \frac{1}{2}\sigma^\mathrm{HF}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{HF}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{HF}\bi{X}\kappa^\mathrm{CC} \\
&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{HF}
  + \sigma^\mathrm{CC}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{HF}
  + \kappa^\mathrm{CC}\zeta^\mathrm{HF}
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{HF} \\
&=
    \frac{1}{2}\sigma^\mathrm{HF}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF} \\
&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{CC}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{CC}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{CC} \\
&+ \sigma^\mathrm{CC}(\bi{Y}\sigma^\mathrm{HF} + \bi{X}\kappa^\mathrm{HF} + \esp^\mathrm{HF}) \\
&+ \kappa^\mathrm{CC}(\bi{W}\kappa^\mathrm{HF} + \bi{X}^\dagger\sigma^\mathrm{HF} + \zeta^\mathrm{HF}) \\
&+ \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{}) \\
&=
    \frac{1}{2}\sigma^\mathrm{HF}\esp^\mathrm{HF}
  + \frac{1}{2}\kappa^\mathrm{HF}\zeta^\mathrm{HF} \\
&+  \frac{1}{2}\sigma^\mathrm{CC}\PCM\sigma^\mathrm{CC}
  + \sigma^\mathrm{CC}\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa^\mathrm{CC}\MM\kappa^\mathrm{CC}
  + \kappa^\mathrm{CC}\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma^\mathrm{CC}\bi{X}\kappa^\mathrm{CC} \\
&+ \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{})
  \end{aligned}
\end{equation}
where we made use of the stationarity condition for the reference
reaction field to eliminate a number of terms. Notice that the
reference reaction energy still appears among the terms in the
functional. However, these terms will not enter in the optimization of
the \acs{CC} wave function and we will therefore drop them in the
following, out of notational convenience:
\begin{equation}
  \begin{aligned}
  U_\mathrm{pol} &=
    \frac{1}{2}\sigma\PCM\sigma
    + \sigma\tilde{\esp}(\tamp{},\tbar{})
  + \frac{1}{2}\kappa\MM\kappa
  + \kappa\tilde{\zeta}(\tamp{},\tbar{})
  + \sigma\bi{X}\kappa
  + \sigma^\mathrm{HF}\tilde{\esp}(\tamp{},\tbar{}) +
  \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{},\tbar{}) \\
  &= \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  + \p^\mathrm{HF}\tilde{\s}(\tamp{},\tbar{})
\end{aligned}
\end{equation}

One can then write the effective Lagrangian as:
\begin{equation}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \sigma) &=
  \braket{\mathrm{HF} | \overline{H_0} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tamp{i} | \overline{H_0} | \mathrm{HF}} \\
  &+ \frac{1}{2}\sigma\PCM\sigma
  + \sigma\tilde{\esp}(\tamp{}, \tbar{})
  + \frac{1}{2}\kappa\MM\kappa
  + \kappa\tilde{\zeta}(\tamp{}, \tbar{})
  + \sigma\bi{X}\kappa
  + \sigma^\mathrm{HF}\tilde{\esp}(\tamp{}, \tbar{})
  + \kappa^\mathrm{HF}\tilde{\zeta}(\tamp{}, \tbar{})
  \end{aligned}
\end{equation}
In accordance with previous work~\cite{Cammi2009-gu, Caricato2011-tx}, the
Hamiltonian $H_0$ is augmented with the terms
$\sigma^\mathrm{HF}\tilde{\esp}(\tamp{}, \tbar{})$ and
$\kappa^\mathrm{HF}\tilde{\zeta}(\tamp{}, \tbar{})$
part appearing in the previous functional yielding the \acl{PTE} (\acs{PTE}) Hamiltonian --
$H$ -- and
eventually the \acl{PTED} (\acs{PTED}) Lagrangian:
\begin{equation}\label{eq:pted-cc}
  \begin{aligned}
  \lag{M}(\tamp{}, \tbar{}, \sigma) &=
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}} \\
  &+ \frac{1}{2}\sigma\PCM\sigma
  + \sigma\tilde{\esp}(\tamp{}, \tbar{})
  + \frac{1}{2}\kappa\MM\kappa
  + \kappa\tilde{\zeta}(\tamp{}, \tbar{})
  + \sigma\bi{X}\kappa \\
  &=
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
  + \frac{1}{2}\p^\dagger\V\p + \p^\dagger\tilde{\s}(\tamp{},\tbar{})
  \end{aligned}
\end{equation}
As noted by \citeauthor{Cammi2009-gu} for the implicit \acs{PCM} model,
the \acs{PTE} Hamiltonian is computed in \acs{MP}-partitioned form by
employing the "solvated" orbitals and Fock matrix:
\begin{equation}
  H = F + \Phi.
\end{equation}
$F$ is a diagonal one-electron operator and $\Phi$ is the
fluctuation potential.
The \acs{PTE} effective Lagrangian ${}^\mathrm{PTE}\lag{M}$ can be
recovered by dropping the classical terms in the Lagrangian
\eqref{eq:pted-cc}:
\begin{equation}
{}^\mathrm{PTE}\lag{M}
  =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
\end{equation}
in accordance with the derivation of \citeauthor{Cammi2009-gu}.
As shown by \citeauthor{Caricato2011-tx}, it is possible to achieve less
drastic approximations of the \acs{PTED} Lagrangian. Such approximations
can be easily derived within the current variational framework. We
postpone this analysis to a later stage of development, \emph{vide
infra}.

\section{Source terms in the classical energy functional}

The variational formulation offers a series of advantages.
The classical energy functional:
\begin{equation}
  U_\mathrm{pol} =
   \frac{1}{2}\sigma\PCM\sigma + \sigma\esp
 + \frac{1}{2}\kappa\MM\kappa + \kappa\zeta
 + \sigma\bi{X}\kappa
\end{equation}
renders itself to a straightforward physical interpretation. In the
three bilinear terms, \ie{} the ones mediated by an interaction
operator, we can identify the self-interaction of the classical
polarization with itself, be it purely implicit:
$\frac{1}{2}\sigma\PCM\sigma$, purely explicit
$\frac{1}{2}\kappa\MM\kappa$ or mixed $\sigma\MMPCM\kappa$.
These terms are positive definite and give rise to an unfavorable
contribution, which is counterbalanced by the linear terms. These terms
mediate the interaction between the induced polarization and the
inducing fields: $\esp$ and $\zeta$. While $\esp$ is quite clearly the
\acl{MEP} (\acs{MEP}), $\zeta$ can either be the molecular electric
field (MMpol and PE models) or again the \acs{MEP} (FQ model).
In any case, both will be determined by the quantum mechanical molecular
charge density and can thus be formulated as expectation values of
one-electron operators. Eventually, this achieves the coupling between
the classical -- $\sigma$ and $\kappa$ -- and
the quantum mechanical variational degrees of freedom.

The \acs{MEP} operator is given as:
\begin{equation}
  \esp = \sum_{PQ}\esp_{PQ}\cons{P}\anni{Q}
\end{equation}
and similarly for the \acs{MM} source operator:
\begin{equation}
  \zeta = \sum_{PQ}\zeta_{PQ}\cons{P}\anni{Q}
\end{equation}
both are nondiagonal in the MO basis. However,
by virtue of the \emph{cluster commutation condition}, their similarity
transformations will truncate after two
terms \cite{Helgaker2000-tz}:
\begin{equation}
  \overline{\esp} = \esp + \tilde{\esp}  = \esp + \commutator{\esp}{T} +
  \frac{1}{2}\commutator{\commutator{\esp}{T}}{T}
\end{equation}
and
\begin{equation}
  \overline{\zeta} = \zeta + \tilde{\zeta}  = \zeta + \commutator{\zeta}{T} +
  \frac{1}{2}\commutator{\commutator{\zeta}{T}}{T}
\end{equation}

In the supermatrix formalism, the similarity-transformed terms will be
written as:
\begin{equation}
  \overline{\s} = \s + \tilde{\s} = \s +
  \commutator{\s}{T} +
  \frac{1}{2}\commutator{\commutator{\s}{T}}{T}
\end{equation}
It is important to single out which terms contribute in the various
expectation values where the similarity-transformed sources will be
involved.
We develop this analysis in terms of excitation ranks of the
operators involved, based on the following:
\begin{equation}
  \braket{\mu_n |
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} | \mathrm{HF}} \neq 0
  \Longleftrightarrow
  n - m_O + k \leq \sum_{i=1}^k n_i \leq n + m_O
\end{equation}
where $n$ is the excitation rank of the $\ket{\mu_n}$ determinant, $k$
is the total number of cluster operators appearing in the nested
commutator, the $n_i$ are their respective excitations ranks and $m_O$
is the \emph{particle rank} of the operator $O$ \cite{Helgaker2000-tz}.
This follows from the realization that:
\begin{equation}
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} \ket{\mathrm{HF}}
\end{equation}
is a linear combination of determinants with excitation ranks $v$ in the
range:
\begin{equation}
  \sum_{i=1}^k n_i - m_O \leq v \leq \sum_{i=1}^k n_i + m_O -k
\end{equation}
The source operator is a nondiagonal, one-electron
operator and hence has particle rank $m_\s = 1$.

Eventually, one can cast the source terms entering the Lagrangian as
follows:
\begin{equation}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \s + \BCHfirst{\s}{T} + \BCHsecond{\esp}{T} | \mathrm{HF}}
\end{equation}

\begin{align}
  \braket{\mathrm{HF} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad 0 \leq t \leq 1
  \Longleftrightarrow t = 1 \\
  \braket{\mathrm{HF} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF}}
  &\quad\quad\text{nonzero contributions} \quad t + u = 1 \\
  \braket{\tbar{i} |
  \esp
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i = 1 \\
  \braket{\tbar{i} |
  \BCHfirst{\s}{T_t}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad i \leq t \leq i + 1 \\
  \braket{\tbar{i} |
  \commutator{\commutator{\s}{T_t}}{T_u}
  | \mathrm{HF} }
  &\quad\quad\text{nonzero contributions} \quad t + u = i + 1
\end{align}

For a general truncation level we accordingly obtain:
\begin{equation}\label{eq:source-term}
  \tilde{\s}(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \BCHfirst{\s}{T} + \BCHsecond{\s}{T} | \mathrm{HF}}
\end{equation}
which shows the unique role the singles amplitudes assume when a
semiclassical environment model is introduced, \emph{vide infra}.
Furthermore, for the \acs{CCS} model ($\mathcal{M} = 1$):
\begin{equation}
  \tilde{\s}^1(\tamp{}, \tbar{}) =
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | \s | \mathrm{HF}}
  + \braket{\tbar{1} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
  + \frac{1}{2}\braket{\tbar{1} | \commutator{\commutator{\s}{T_1}}{T_1} | \mathrm{HF}}
\end{equation}
while for the \acs{CCSD} model ($\mathcal{M} = 2$):
\begin{equation}
  \tilde{\s}^2(\tamp{}, \tbar{}) =
  \tilde{\s}^1(\tamp{}, \tbar{})
  + \braket{\tbar{1} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \BCHfirst{\s}{T_2} | \mathrm{HF}}
  + \braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}
\end{equation}
Eventually, within the \acs{CCSDT} model ($\mathcal{M} = 3$) four more
terms are added:
\begin{equation}
  \begin{aligned}
  \tilde{\s}^3(\tamp{}, \tbar{}) &=
  \tilde{\s}^2(\tamp{}, \tbar{})
  + \braket{\tbar{2} | \BCHfirst{\s}{T_3} | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \BCHfirst{\s}{T_3} | \mathrm{HF}}
  + \frac{1}{2}
  \braket{\tbar{3} | \commutator{\commutator{\s}{T_2}}{T_2} |
  \mathrm{HF}}
  + \braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} |
  \mathrm{HF}}
  \end{aligned}
\end{equation}
We note that:
\begin{itemize}
    \item double commutators involving the highest order cluster
      operators can only lead to nonzero terms when $T_1$ is also
      involved and by projection onto the highest order left \acs{CC}
      state, \emph{e.g.}
      $\braket{\tbar{2} | \commutator{\commutator{\s}{T_1}}{T_2} | \mathrm{HF}}$
      and
      $\braket{\tbar{3} | \commutator{\commutator{\s}{T_1}}{T_3} | \mathrm{HF}}$
\end{itemize}

\section{The PTED-CC equations}\label{sec:pted-cc-equations}

Differentiation of the effective Lagrangian in Eq.~\eqref{eq:pted-cc}
with respect to the variational parameters $\tamp{\mu_i}$,
$\tbar{\mu_i}$, $\sigma$ and $\kappa$ yields the \acs{PTED}-\acs{CC}
equations:
\begin{equation}
  \begin{aligned}
   \pderiv{\lag{M}}{\tbar{\mu_i}} &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
    + \sigma\braket{\mu_i | \overline{\esp} | \mathrm{HF}}
    + \kappa\braket{\mu_i | \overline{\zeta} | \mathrm{HF}}
    = \tampEq{\mu_i}(\tamp{}, \tbar{}, \sigma, \kappa) \\
    \pderiv{\lag{M}}{\tamp{\mu_i}} &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} \\
    &+ \sigma\lbrace
    \braket{\mathrm{HF} | \commutator{\esp}{\cluster{\mu_i}} +
    \commutator{\commutator{\esp}{T}}{\cluster{\mu_i}}| \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\esp}}{\cluster{\mu_i}} | \mathrm{HF}}
             \rbrace
             \\
    &+ \kappa\lbrace
    \braket{\mathrm{HF} | \commutator{\zeta}{\cluster{\mu_i}} +
    \commutator{\commutator{\zeta}{T}}{\cluster{\mu_i}}| \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\zeta}}{\cluster{\mu_i}} | \mathrm{HF}}
             \rbrace
    = \tbarEq{\mu_i}(\tamp{}, \tbar{}, \sigma, \kappa) \\
    \pderiv{\lag{M}}{\sigma} &= \PCM\sigma + \MMPCM\kappa
    + \tilde{\esp}(\tamp{},\tbar{}) =
    \Omega_\sigma(\tamp{}, \tbar{}, \sigma, \kappa) \\
    \pderiv{\lag{M}}{\kappa} &= \MM\kappa + \MMPCM^\dagger\sigma + \tilde{\zeta}(\tamp{},\tbar{}) =
    \Omega_\kappa(\tamp{}, \tbar{}, \sigma, \kappa)
  \end{aligned}
\end{equation}
Given the evident similarities between the implicit and explicit terms
in the amplitudes and multipliers equations, we adopt the supermatrix
formalism for the classical polarizable terms. The equations are thus
recast as:
\begin{equation}
  \begin{aligned}
   \pderiv{\lag{M}}{\tbar{\mu_i}} &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
   + \p^\dagger\braket{\mu_i | \overline{\s} | \mathrm{HF}}
    = \tampEq{\mu_i}(\tamp{}, \tbar{}, \p) \\
    \pderiv{\lag{M}}{\tamp{\mu_i}} &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} \\
    &+ \p^\dagger\lbrace
    \braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_i}} +
    \commutator{\commutator{\s}{T}}{\cluster{\mu_i}}| \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} | \commutator{\overline{\s}}{\cluster{\mu_i}} | \mathrm{HF}}
             \rbrace
    = \tbarEq{\mu_i}(\tamp{}, \tbar{}, \p) \\
    \pderiv{\lag{M}}{{\p^{\dagger}}} &=
    \V\p + \tilde{\s}(\tamp{}, \tbar{}) =
    \Omega_\p(\tamp{}, \tbar{}, \p)
  \end{aligned}
\end{equation}

\section{Approximate Effective Lagrangians}\label{sec:approximate-lagrangians}

We have already noted that the leading terms in the correlated source operator expectation
value, Eq.~\eqref{eq:source-term}, prominently involve the singles
cluster operator and the singles left \acs{CC} state.
This simple fact led \citeauthor{Caricato2011-tx} to propose two
intermediate approximations in between the \acs{PTE} and \acs{PTED}
schemes: the \acs{PTES} and \acs{PTE(S)} schemes.
The \acs{PTE} scheme is advantageous since no coupling between the
\acs{CC} amplitude and multiplier equations is introduced. However, the
polarization included in the correlation treatment is the same as for
the reference determinant, an approximation which appears questionable
from the point of view of classical electrostatics. On the contrary, the
\acs{PTED} scheme describes the fully equilibrated electrostatic
picture, at the expense of coupling the amplitude and multiplier
equations.
However, if we truncate the correlated source operator expectation value
to the first term in Eq.~\eqref{eq:source-term}, the coupling between
the \acs{CC} right and left states is broken, while still improving upon the
reference reaction field with correlated contributions.
This is the essence of the \acs{PTE(S)} approximation. The effective
Lagrangian would then read as:
\begin{equation}\label{eq:ptes-cc}
  ^{\mathrm{PTE(S)}}\lag{M}(\tamp{}, \tbar{}, \p) =
  \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{H} | \mathrm{HF}}
  + \frac{1}{2}\p^\dagger\V\p + \p^\dagger
  \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}
\end{equation}
which uncouples all equations:
\begin{equation}
  \begin{aligned}
   \pderiv{\lag{M}}{\tbar{\mu_i}} &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
   =
   ^{\mathrm{PTE(S)}}\tampEq{\mu_i}(\tamp{}, \tbar{}, \p) \\
    \pderiv{\lag{M}}{\tamp{\mu_1}} &=
    \denom{\mu_1}\tbar{\mu_1} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}}
    + \p^\dagger
    \braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_1}} | \mathrm{HF}}
    =
    ^{\mathrm{PTE(S)}}\tbarEq{\mu_1}(\tamp{}, \tbar{}, \p) \\
    \pderiv{\lag{M}}{\tamp{\mu_i}} &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}}
    =
    ^{\mathrm{PTE(S)}}\tbarEq{\mu_i}(\tamp{}, \tbar{}, \p) \quad \forall i \neq 1 \\
    \pderiv{\lag{M}}{{\p^{\dagger}}} &=
    \V\p + \braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}} =
    ^{\mathrm{PTE(S)}}\Omega_\p(\tamp{}, \tbar{}, \p)
  \end{aligned}
\end{equation}
The amplitude equations are formally equivalent to those derived in the
\acs{PTE} scheme. The equations for multipliers are also formally
equivalent to the \acs{PTE} equations, apart from the equation
determining the singles multipliers which is augmented by an additional term:
$\p^\dagger\braket{\mathrm{HF} | \commutator{\s}{\cluster{\mu_1}} |
\mathrm{HF}}$
This term is not, however, coupled to the other equations given that
$\p$ is fixed once the singles amplitudes have been determined.
Practical implementation of the \acs{PTE(S)} scheme will solve the
\acs{PTE}-like amplitude equation, form the source term
$\braket{\mathrm{HF} | \BCHfirst{\s}{T_1} | \mathrm{HF}}$, solve for
the polarization degrees of freedom and calculate the polarization
energy as a correction to the \acs{CC} energy.
Only when a first-order molecular property, such as the molecular
gradient, is needed will the multipliers equations need to be solved.
